---
title: "Econometric Analysis: Carbon Pricing and Electricity Demand"
author: "Adinda Septiani - 33917949"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(janitor)
library(fixest)
library(modelsummary)
library(kableExtra)
library(lubridate)  
```

---

# 1. Declaration of Analytical Ambition

This analysis is **descriptive**. It estimates the conditional association between
coal intensity and electricity demand following the introduction of Australian carbon
pricing in July 2012, using a balanced state-level panel. No claim is made that the
estimated relationships are causal. Because all five National Electricity Market (NEM)
states were subject to the same policy simultaneously, there is no untreated control
group, and a standard difference-in-differences identification strategy is not available.
The coefficients reported below should be interpreted as conditional correlations for what
is associated with demand, after absorbing fixed differences across states and common
time trends and not as treatment effects.

---

# 2. Load Data

```{r load-data}
candidate_paths <- c("data/clean/merged_data.csv")
data_path <- candidate_paths[file.exists(candidate_paths)][1]

if (is.na(data_path)) {
  stop("No dataset found. Expected data/clean/merged_data.csv")
}

analysis_data <- read_csv(data_path) |>
  clean_names() |>
  mutate(
    post_jul2012 = if_else(year > 2012 | (year == 2012 & month >= 7), 1L, 0L),
    log_demand   = log(avg_demand),
    log_price    = log(avg_price),
    log_pop      = log(population),
    avg_temp_sq = avg_temp^2, # Quadratic term for temperature to capture non-linear effects
    # Quarter fixed effect label for time FE
    time_id      = paste(year, sprintf("%02d", month), sep = "-"),
    # Interaction: post-policy × coal share (continuous treatment intensity)
    post_x_coal  = post_jul2012 * coal_share
  )

glimpse(analysis_data)
```

The cleaned dataset contains `r nrow(analysis_data)` observations across
`r n_distinct(analysis_data$region)` states and
`r n_distinct(analysis_data$time_id)` quarterly time periods (2010–2014).
The panel is balanced: each state contributes exactly 20 observations.

---

# 3. Econometric Specification

## 3.1 Functional Form and Regressors

The outcome variable is the log of average quarterly electricity demand, 
$\log(\text{demand}_{it})$. In the EDA, the distribution of avg_demand is multi-modal and 
right-skewed, with observations clustering at very different levels across states 
(Section 3.4.1). NSW sit at 8,000–9,500 MWh, while TAS and SA cluster around 1,000–1,500 MWh. 
That's nearly a 10x difference in raw units. In a linear model, a one-unit change in a 
regressor would imply the same absolute change in MWh regardless of whether you're in NSW or 
TAS. That's economically implausible. For example, a $1 price increase means something very 
different to a large state vs. a small one. Taking logs means coefficients become 
proportional/percentage effects, which makes much more sense when states vary so dramatically 
in baseline size. 

The main equation estimated is:

$$
\log(\text{demand}_{it}) = \alpha_i + \lambda_t + \beta_1 \, \text{post}_{t}
  + \beta_2 \, (\text{post}_{t} \times \text{coalshare}_{i})
  + \gamma_2 \, \text{temp}_{it}
  + \gamma_2 \, \text{temp}_{it}^2
  + \gamma_3 \log(\text{pop}_{it})
  + \varepsilon_{it}
$$

where:

- $i$ indexes state, $t$ indexes quarter-year.
- $\alpha_i$ are **state fixed effects**, absorbing all time-invariant state-level
  differences (e.g., industrial structure, climate zone).
- $\lambda_t$ are **time fixed effects**, absorbing common shocks in each period
  (e.g., national business cycle, energy market conditions).
- $\text{post}_{t}$ is a binary indicator equal to 1 from July 2012 onward.
- $\text{post}_{t} \times \text{coalshare}_{i}$ is the key variable of interest.
  It captures whether the post-policy change in demand was systematically larger
  in states with a higher share of coal in electricity generation.
- $\text{temp}_{it}$, $\text{temp}_{it}^2$, and $\log(\text{pop}_{it})$ are
  demand-side controls for seasonal temperature and state size.

### Justification of regressors:

- *Coal share* varies primarily across states (with modest within-state time variation),
  making it a proxy for carbon intensity exposure. It bounded between 0 and 1, states like 
  NSW and VIC have coal shares above 0.80 while Tasmania is at zero. Keeping it in its 
  natural 0–1 scale makes the interaction coefficient directly interpretable.
- *Temperature* is included because the EDA showed clear seasonality in demand. Without
  it, the time fixed effects at the quarter level would need to fully absorb seasonal
  variation. It is meaningful in linear scales, a one-degree change has a consistent 
  real-world interpretation. Temperature is included as a quadratic ($\text{temp}_{it}$, 
  $\text{temp}_{it}^2$) because Figure 6.2 of the EDA shows a U-shaped within-state 
  relationship between temperature and demand, consistent with dual heating and cooling 
  demand. A linear term would be misspecified.
- *Population* controls for scale differences across states. Similar with *Demand*, it is 
  log-transformed because its distribution is clustering and has the large scale differences 
  across states. It also has high correlation with demand (Section 7 EDA), making Log-Log
  more natural for explaining the relationship.
- *Price* is excluded as a demand-side control because it likely acts as a mediator between 
  carbon pricing and demand. Carbon pricing likely raised electricity prices, which then 
  reduced demand, meaning controlling for price partially absorbs the association of interest.
  Endogeneity concerns around price are also more severe, as price is determined 
  simultaneously with demand in the market. Including it would therefore complicate the i
  nterpretation of the coefficients and potentially bias the estimates of the association 
  between coal share and demand.

## 3.2 Sample

Five NEM states (NSW, QLD, SA, TAS, VIC), quarterly observations from Q1 2010 to Q4 2014,
giving N = 100 observations. Western Australia and Northern Territory are excluded because 
they are not part of the NEM. The five NEM states are retained because they share the same 
30-minute dispatch and settlement system, making their demand and price data directly 
comparable. This also ensures the panel remains balanced. Meanwhile, the time window is 
chosen symmetrically around the carbon pricing period. Australia's carbon price was 
introduced in July 2012 and repealed in July 2014 by the Abbott Coalition government, which 
argued it imposed an economic burden. This repeal creates a natural upper bound for the 
sample. The pre-policy window is set to match,starting from Q1 2010 gives approximately 2.5 
years before the policy (10 quarters) and 2.5 years under it (10 quarters), providing a 
symmetric comparison that supports visual inspection of pre-policy trends.

## 3.3 Error Structure

Standard errors are clustered at the **state level** to allow for arbitrary serial
correlation within states over time. The error structure must account for the fact that 
observations within the same state are not independent across time. The standard assumption 
of independent errors across all observations is potentially violated as the shocks that 
happen within states is likely to persist in consecutive quarter. Clustering allows the 
variance-covariance matrix to be fully unrestricted within each state across all time periods, 
giving valid inference under arbitrary serial correlation and heteroskedasticity within 
clusters. One limitation of this approach is that cluster-robust inference is justified as 
the number of clusters grows large. With only five states, this asymptotic approximation is 
poor, meaning significance results should be interpreted with caution.

---

# 4. Results

```{r regressions}
# Specification 1: Pooled OLS, no fixed effects
m1 <- feols(
  log_demand ~ post_jul2012 + post_x_coal + avg_temp + avg_temp_sq + log_pop,
  data    = analysis_data,
  vcov    = "hetero"
)

# Specification 2: State fixed effects only
m2 <- feols(
  log_demand ~ post_jul2012 + post_x_coal + avg_temp + avg_temp_sq + log_pop
             | region,
  data    = analysis_data,
  cluster = ~region
)

# Specification 3: State + time fixed effects (preferred)
m3 <- feols(
  log_demand ~ post_jul2012 + post_x_coal + avg_temp + avg_temp_sq + log_pop
             | region + time_id,
  data    = analysis_data,
  cluster = ~region
)
```

```{r regression-table}
modelsummary(
  list(
    "OLS"              = m1,
    "State FE"         = m2,
    "State + Time FE"  = m3
  ),
  coef_rename = c(
    "post_jul2012" = "Post July 2012",
    "post_x_coal"  = "Post × Coal Share",
    "avg_temp"     = "Temperature",
    "avg_temp_sq"  = "Temperature Squared",
    "log_pop"      = "Log(Population)"
  ),
  stars    = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  gof_map  = c("nobs", "r.squared", "adj.r.squared"),
  title    = "Table 1: Conditional Association Between Coal Intensity and Electricity Demand",
  notes    = "Dependent variable: log(avg quarterly electricity demand). Standard errors in
              parentheses: heteroskedasticity-robust for OLS; clustered by state for
              specifications (2) and (3). *** p<0.01, ** p<0.05, * p<0.1.",
  output   = "kableExtra"
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  add_header_above(c(" " = 1, "Pooled" = 1, "Within-State" = 1, "Within-State & Time" = 1))
```

---

# 5. Interpretation of Main Coefficients

The preferred specification is Column (3), which includes both state and time fixed
effects and thereforeidentifies associations from variation that remains after removing 
persistent state-level differences and common time trends. Column (1) is a naive OLS 
regression that does not control for any fixed effects. (2) controls for state fixed effects, 
which absorb all time-invariant differences across states (e.g., industrial structure, 
climate), but it does not control for common time shocks that affect all states 
simultaneously. The inclusion of fixed effects in Column (3) means that the coefficients 
should be interpreted as follows:

**Post × Coal Share** ($\hat{\beta}_2$): This is the coefficient of primary interest.
A negative value indicates that demand fell more in coal-intensive states after the policy, 
relative to lower-coal states, conditional all controls and on fixed effects. Moving from 
pooled OLS to the preferred specification with state and time fixed effects, the coefficient 
on Post × Coal Share changes from 0.145 (SE = 0.138)to -0.06 (SE = 0.054), suggesting that 
some of the raw association was driven by state-level heterogeneity and common time trends, 
which the fixed effects absorb. The magnitude of coefficient is -0.06, meaning that a 
one-unit increase in coal share is associated with approximately 6% lower electricity 
demand in the post-policy period relative to the pre-policy period, holding state fixed 
effects, time effects, temperature, and population constant. While showing that higher coal 
share is associated with lower demand after the policy, the effect is not statistically 
significant at conventional levels (p > 0.1). This association should not be interpreted as a 
causal effect of carbon pricing on demand, due to potential confounding factors that vary 
across states and over time.

**Post July 2012** ($\hat{\beta}_1$): This is the average change in log demand across all 
states after the policy introduction, conditional on all controls and fixed effects. A 
negative value would suggest that demand declined on average across all states after the 
policy, even after controlling for temperature and population. This coefficient does not 
appear in the preferred specification with time fixed effects, because the time fixed effects at the quarter level 
absorb all common time shocks, including any average post-policy shift. In the OLS and state 
FE specifications, the coefficient is -0.147 (SE = 0.108) and -0.011 (SE = 0.020)
respectively, but it is not statistically significant in either case.

**Temperature**: Temperature enters the model as a quadratic — both the linear term 
(−0.059, SE = 0.014) and the squared term (0.002, SE = 0.000) are included to capture the 
U-shaped relationship identified in Figure 6.2 EDA. The two coefficients cannot be 
interpreted independently, together they describe how demand changes across the temperature 
range. The turning point, where demand reaches its minimum, is calculated as 
$$-\frac{\hat{\gamma}_2}{2\hat{\gamma}_3} = -\frac{(-0.059)}{2 \times 0.002} \approx 14.75°C$$
Below this point, colder temperatures are associated with higher demand driven by heating load and hotter 
temperatures are associated with higher demand driven by cooling load. Both the linear and 
squared terms are statistically significant. The linear term at the 5% level and the squared 
term at the 1% level, confirming that the non-linear specification is appropriate and that 
temperature is a meaningful predictor of electricity demand after controlling for state and 
time fixed effects. As with all results in this analysis, this should be interpreted as a 
conditional association rather than a causal effect.

**Log(Population)**: This is the average change in log demand associated with a 1% increase
in population, conditional on all other factors. A positive coefficient would suggest that
larger states have higher demand, which is expected. The coefficient moves from 0.778 (SE =
0.034) in pooled OLS to 0.180 (SE = 0.020) in the preferred specification, meaning that a 1% 
increase in population is associated with approximately a 0.180% increase in electricity 
demand, after controlling for state and time fixed effects, temperature, and the post-policy 
period. This effect is not statistically significant in the preferred specification (p > 0.1), 
which likely because population barely moves within a state over this short window, so the 
fixed effects leave almost no within-state variation for population to explain.

> **Note:** R squared is 0.999 in Column (3), mechanically high due to state and time fixed 
effects absorbing most raw variation, as discussed. The within R squared is the more 
meaningful fit measure here. The improvement from 0.917 (pooled OLS) to 0.999 (state + time 
FE) reflects how much of the variation in demand is explained by persistent state differences 
and common time trends rather than the regressors themselves.

---

# 6. Threats and Limitations

Because this analysis is declared **descriptive**, the threats section focuses on
limitations to the conditional correlations reported, rather than threats to causal
identification.

## 6.1 Omitted Confounders

The most important limitation is that the `post × coal share` association may reflect
factors other than the carbon price. For example:

- **Energy efficiency programs**: Several states introduced energy efficiency initiatives 
  (e.g., Greenhouse and Energy Minimum Standards Act 2012, lower-emission technologies) 
  around the same period. If these were more prevalent in high-coal states, the observed 
  demand association would be at least partly driven by efficiency rather than carbon 
  pricing. This would bias the association downward.
- **Industrial composition**: If coal-intensive states also had more energy-intensive
  industries that declined over this period for unrelated reasons (e.g., manufacturing
  contraction), the negative association could reflect structural deindustrialisation
  rather than a price-driven demand response. Again, this would bias the association downward.

State fixed effects absorb *time-invariant* state-level characteristics, but they do not
absorb state-specific *time-varying* confounders. This is the primary limitation of the
descriptive interpretation.

## 6.2 Sample Selection

The sample covers only five NEM states over five years. The results may not generalise beyond 
the NEM (e.g., Western Australia which has major fuel extraction and mining sector). 
Observations without population data were dropped during the cleaning process, as population 
is a required control variable in the specification. The short time window (2010–2014) limits 
the ability to separate policy effects from medium-run structural trends such as the ongoing 
shift away from coal in Australia's electricity mix, especially when demand of electricity is 
inelastic in the short run. A longer time series would allow for better understanding of how 
demand evolved before and after the policy, and whether the observed association is part of 
a longer-term trend rather than a discrete shift around the policy introduction.

## 6.3 Measurement Limitations

- **Coal share**: Coal sharechanges very little within each state over time. The mix of power plants in 
  a state is largely fixed by existing infrastructure and changes slowly over decades. 
  Because state fixed effects absorb everything that is constant within a state, the 
  interaction term post × coal share is identified from only the small year-to-year movements 
  in coal share within each state. This means the estimate is based on limited variation and 
  should be interpreted cautiously.
- **Quarterly aggregation**: Demand and price data in the NEM are originally recorded every 
  30 minutes. They were aggregated to quarterly averages in this analysis due to population 
  data only being available at the quarterly level. This means short-run fluctuations in 
  demand are smoothed out, and the analysis captures longer-run average patterns rather than 
  immediate responses to price or policy changes within a quarter.

## 6.4 Small Number of Clusters

Standard errors are clustered at the state level to account for correlation within states 
over time. However, this approach works best when there are many clusters which ideally 30 or 
more. With only five states, the clustered standard errors may not be fully reliable, and 
statistical significance results should be interpreted with caution. This is an inherent 
limitation of working with state-level data where the number of units is fixed by geography.

---

# Conclusion

This analysis estimates the conditional association between coal intensity and electricity
demand in Australian NEM states before and after the introduction of carbon pricing in
July 2012. Using a panel with state and time fixed effects, the `Post × Coal Share`
coefficient captures whether demand trajectories differed systematically across states
with different carbon intensities in the post-policy period, after absorbing common
time shocks and persistent state-level differences.

The preferred specification finds a small negative association (−0.060, SE = 0.054)
between coal intensity and post-policy demand, suggesting that higher coal-share states
were associated with slightly lower demand after the policy introduction. However, this
association is not statistically significant at conventional levels, and the absence of
a large or significant demand response is itself a meaningful finding. Electricity demand
is largely inelastic in the short run. Households and businesses do not rapidly change
their consumption patterns in response to price signals, and the energy mix underlying
electricity generation cannot shift immediately. The transition away from coal-intensive
generation is a structural process that unfolds over decades, not quarters. Together,
these factors suggest that detecting a strong differential demand response over a
two-and-a-half year window would be unlikely even if the carbon price had meaningful
long-run effects.

The one variable that does show a statistically significant association is temperature,
whose quadratic specification confirms a U-shaped relationship between temperature and
demand, which areconsistent with dual heating and cooling load across seasons. This 
highlights that short-run demand variation within states is driven primarily by seasonal 
factors rather than policy-induced price changes over this window.

The analysis is descriptive: the panel structure and absence of a clean control group
preclude a credible causal claim. All coefficients should be interpreted as conditional
associations rather than causal effects. The results nonetheless provide suggestive
evidence on the correlates of electricity demand in the context of carbon pricing, and
serve as a foundation for future work with longer time series, higher-frequency data,
or alternative identification strategies that could better isolate the policy effect
from structural trends in Australia's electricity market.
