---
title: "Robustness Analysis: Carbon Pricing and Electricity Demand"
author: "Adinda Septiani - 33917949"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(janitor)
library(fixest)
library(modelsummary)
library(kableExtra)
library(lubridate)
```

---

# 1. Main Result and Declaration

This analysis is **descriptive**. The primary analysis estimated the conditional
association between coal intensity and electricity demand in Australian NEM states
before and after the introduction of carbon pricing in July 2012. All results should
be interpreted as conditional correlations, not causal effects.

The main result from the preferred specification (state + time fixed effects) is:

- **Post × Coal Share = −0.060 (SE = 0.054)**: States with higher coal share are
  associated with approximately 6% lower electricity demand in the post-policy period
  relative to the pre-policy period, after controlling for state fixed effects, time
  fixed effects, temperature (quadratic), and population. This association is **not
  statistically significant** at conventional levels (p > 0.1).

- **Temperature** enters as a quadratic and is statistically significant, confirming
  a U-shaped heating-cooling relationship with demand. This signals the short-term 
  demand variation within states is associated with seasonal temperature changes rather 
  than policy-related factors.

The robustness checks below stress-test whether the sign, magnitude, and significance
of the `Post × Coal Share` coefficient are sensitive to alternative control sets,
samples, functional forms, and inference approaches.

---

# 2. Data

```{r load-data}
candidate_paths <- c("data/clean/merged_data.csv")
data_path <- candidate_paths[file.exists(candidate_paths)][1]

if (is.na(data_path)) {
  stop("No dataset found. Expected data/clean/merged_data.csv")
}

analysis_data <- read_csv(data_path) |>
  clean_names() |>
  mutate(
    post_jul2012  = if_else(year > 2012 | (year == 2012 & month >= 7), 1L, 0L),
    log_demand    = log(avg_demand),
    log_pop       = log(population),
    avg_temp_sq   = avg_temp^2,
    time_id       = paste(year, sprintf("%02d", month), sep = "-"),
    post_x_coal   = post_jul2012 * coal_share
  )
```

---

# 3. Robustness Checks

## 3.1 Alternative Control Sets

Three specifications test whether the main result is sensitive to the set of
controls included:

- **No controls**: Only `post × coal share` and fixed effects — no temperature,
  no population. Tests whether controls are driving or masking the association.
- **Preferred**: The main specification with temperature (quadratic) and
  log(population).
- **Additional controls**: Adds `coal_share` as a standalone level term alongside
  the interaction, to check whether the interaction coefficient is sensitive to
  directly conditioning on the main effect of coal share.

```{r control-sets}
# No controls (interaction + FE only)
r_nocontrols <- feols(
  log_demand ~ post_x_coal | region + time_id,
  data    = analysis_data,
  cluster = ~region
)
summary (r_nocontrols)

# Preferred specification
r_preferred <- feols(
  log_demand ~ post_x_coal + avg_temp + avg_temp_sq + log_pop
             | region + time_id,
  data    = analysis_data,
  cluster = ~region
)
summary(r_preferred)

# Additional coal_share level term
r_addcontrols <- feols(
  log_demand ~ post_x_coal + coal_share + avg_temp + avg_temp_sq + log_pop
             | region + time_id,
  data    = analysis_data,
  cluster = ~region
)
summary(r_addcontrols)
```

---

## 3.2 Alternative Samples

Two sample restrictions test whether the result is driven by a specific state or
time window:

- **Drop TAS**: Tasmania has a coal share of exactly 0 throughout the sample,
  making it an extreme point in the coal share distribution. Dropping it checks
  whether the interaction is identified primarily from TAS vs. the rest.
- **Narrow window (2011–2013)**: Restricts to 1.5 years around the policy
  introduction to reduce the influence of longer-run structural trends unrelated
  to the carbon price.

```{r alt-samples}
# Drop Tasmania
data_notas <- analysis_data |> filter(region != "TAS")

r_notas <- feols(
  log_demand ~ post_x_coal + avg_temp + avg_temp_sq + log_pop
             | region + time_id,
  data    = data_notas,
  cluster = ~region
)
summary(r_notas)

# Time window 2011-2013 only
data_narrow <- analysis_data |>
  filter(year >= 2011 & year <= 2013)

r_narrow <- feols(
  log_demand ~ post_x_coal + avg_temp + avg_temp_sq + log_pop
             | region + time_id,
  data    = data_narrow,
  cluster = ~region
)
summary(r_narrow)
```

---

## 3.3 Alternative Functional Form

Tests whether the log transformation of the outcome is driving the result by
running the preferred specification in **levels** (avg_demand, not logged):

```{r alt-form}
# Levels outcome instead of log
r_levels <- feols(
  avg_demand ~ post_x_coal + avg_temp + avg_temp_sq + log_pop
             | region + time_id,
  data    = analysis_data,
  cluster = ~region
)
summary(r_levels)
```

---

## 3.4 Alternative Inference

Tests whether the main result is sensitive to the choice of standard errors.
With only 5 clusters, cluster-robust SEs may understate the true uncertainty.
HC3 heteroskedasticity-robust SEs provide an alternative that does not rely
on the small-cluster asymptotic approximation:

```{r alt-inference}
# HC3 robust SEs instead of clustered
r_hc3 <- feols(
  log_demand ~ post_x_coal + avg_temp + avg_temp_sq + log_pop
             | region + time_id,
  data = analysis_data,
  vcov = "HC3"
)
summary(r_hc3)

# Add coal share level term + HC3 robust SEs
r_coal_hc3 <- feols(
  log_demand ~ post_x_coal + coal_share+ avg_temp + avg_temp_sq + log_pop
             | region + time_id,
  data = analysis_data,
  vcov = "HC3"
)
summary(r_coal_hc3)
```

---

## 3.5 Placebo Time Cutoff 

Since this is a descriptive analysis, a key concern is whether the conditional
correlation is driven by a pre-existing trend rather than anything specific to
the post-July 2012 period. A placebo test replaces the true policy cutoff
(July 2012) with a fake cutoff one year earlier (July 2011), using only
pre-policy data (2010–2012). If `placebo × coal share` is also negative and
significant, it suggests the result reflects a pre-existing trend unrelated
to the policy period:

```{r placebo}
# Pre-policy data only with fake July 2011 cutoff
data_placebo <- analysis_data |>
  filter(year <= 2012) |>
  mutate(
    placebo_post   = if_else(year > 2011 | (year == 2011 & month >= 7), 1L, 0L),
    placebo_x_coal = placebo_post * coal_share,
    time_id_p      = paste(year, sprintf("%02d", month), sep = "-")
  )

r_placebo <- feols(
  log_demand ~ placebo_x_coal + avg_temp + avg_temp_sq + log_pop
             | region + time_id_p,
  data    = data_placebo,
  cluster = ~region
)
summary(r_placebo)
```

---

# Robustness Table

```{r robustness-table}
modelsummary(
  list(
    "Preferred"       = r_preferred,
    "No Controls"     = r_nocontrols,
    "Add Coal Level"  = r_addcontrols,
    "Drop TAS"        = r_notas,
    "2011-2013"       = r_narrow,
    "Levels"          = r_levels,
    "HC3 SE"          = r_hc3,
    "Add Coal Level"  = r_coal_hc3,
    "Placebo"         = r_placebo
  ),
  coef_map = c(
    "post_x_coal"    = "Post × Coal Share",
    "placebo_x_coal" = "Placebo × Coal Share",
    "coal_share"     = "Coal Share (level)",
    "avg_temp"       = "Temperature",
    "avg_temp_sq"    = "Temperature²",
    "log_pop"        = "Log(Population)"
  ),
  stars   = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  gof_map = c("nobs", "r.squared"),
  title   = "Table 2: Robustness Checks — Post × Coal Share Coefficient",
  notes   = "Dependent variable: log(avg quarterly electricity demand) unless noted.
             Columns (1)–(3) and (6)–(8) use 2010–2014 full sample. Column (2) has no controls. 
             Column (3) include coal share as a control. Column (4) excludes 
             Tasmania. Column (5) restricts to 2011–2013. Column (6) uses demand in levels (MWh). 
             Column (7) uses HC3 robust SEs instead of clustered. Column (8) use HC3 robust SEs for 
             column (3) specification.Column (9) uses pre-policy data only
             (2010–2012) with a placebo cutoff at July 2011. All specifications include
             state and time fixed effects. Standard errors clustered by state except
             Column (7). *** p<0.01, ** p<0.05, * p<0.1.",
  output  = "kableExtra"
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  add_header_above(c(
    " "             = 1,
    "Main"          = 1,
    "Alt. Controls" = 2,
    "Alt. Sample"   = 2,
    "Alt. Form"     = 1,
    "Alt. SE"       = 2,
    "Placebo"       = 1
  ))
```
``` {r coefficient plot}
library(ggplot2)

# Manually create a dataframe of your robustness results
coef_data <- data.frame(
  spec = c(
    "(1) Preferred",
    "(2) No Controls",
    "(3) Add Coal Level",
    "(4) Drop TAS",
    "(5) 2011-2013",
    "(6) HC3 SE",
    "(7) Add Coal Level + HC3",
    "(8) Placebo"
  ),
  estimate = c(-0.060, -0.033, -0.098, -0.052, -0.085, -0.060, -0.098, -0.011),
  se       = c(0.054,  0.039,  0.045,  0.030,  0.044,  0.038, 0.043, 0.025)
) |>
  mutate(
    ci_low  = estimate - 1.96 * se,
    ci_high = estimate + 1.96 * se,
    # Keep order as listed above
    spec = factor(spec, levels = rev(spec))
  )

ggplot(coef_data, aes(x = estimate, y = spec)) +
  geom_vline(xintercept = -0.060, linetype = "dashed",
             color = "grey50", linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "solid",
             color = "red", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high),
                 height = 0.2, color = "steelblue") +
  geom_point(color = "steelblue", size = 3) +
  labs(
    title = "Figure 1. Post × Coal Share Coefficient Across Specifications",
    x     = expression(hat(beta)["Post × Coal Share"]),
    y     = NULL,
    caption = "Error bars show 95% confidence intervals. 
    Dashed line marks the preferred specification estimate (−0.060). 
    Red line marks zero.
    Error bar crossing zero indicates insignificance at 10% level."
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.margin = margin(10, 20, 10, 10)
  ) +
coord_cartesian(clip = "off")
```


---

# 4. Interpretation of Robustness Checks

## 4.1 Alternative control sets (Columns 1–3)

**Did the result hold?** Report whether the `Post × Coal Share` coefficient
stays negative and its significance changes across columns 1–3.

The no-controls specification (Column 2) removes temperature and population resulting in 
the coefficient changing to  -0.033 (SE = 0.039) and still insignificant. This suggests 
that the controls are not inflating the negative association, but they are also not 
masking a significant relationship. They rather help to isolate the association. The 
additional controls specification (Column 3) adds the coal share level term increase the 
coefficient magnitude to -0.098 (SE = 0.045) and makes it significant at the 10% level. 
Before adding the coal share term, the model treats coal share as time-invariant and 
absorbs it into the state fixed effects. But coal share actually does change slightly 
within states over time. Because it's absorbed into the FE, the model cannot use those 
small within-state changes to precisely estimate how coal share affects the post-policy 
demand response. When we add coal share as its own variable, the model can use both the 
within-state variation in coal share over time and the cross-state differences to 
estimate the post-policy association more precisely, which is why it becomes significant.
The coal share level term itself is significant at 1% level, indicating that states with 
higher coal share have lower demand on average, which is expected given coal-intensive 
states tend to be more populated. Therefore, the preferred specification's insignificance 
is partly a consequence of near-collinearity between coal share and state fixed effects 
rather than a true absence of association.

## 4.2 Alternative samples (Columns 4–5)

**Did the result hold?** Compare the coefficient in the drop-TAS and narrow
window specifications to the preferred.

Tasmania's coal share of exactly 0 makes it the anchor of the lower end of
the coal share distribution. If dropping TAS substantially changes the
coefficient, it means the interaction is being identified primarily from the
contrast between TAS and the rest rather than from variation across all states.
Dropping Tasmania does not substantially change the coefficient to -0.052 (SE = 0.030).
The standard error decreases but the result remain insignificant, which suggests the 
result is not being driven solely by the contrast between TAS and the other states.
Additionally, the coefficient of log(population) increase to 1.783 from 0.180 in the 
preferred specification. This suggests that including TAS, which has a small population, 
may have been distorting the population coefficient, although it remians insignificant in 
both specifications.

The narrow window check addresses whether the association is driven by events
in the early or late parts of the sample rather than the period immediately
around the policy. Restricting to 2011–2013 gives a coefficient of -0.085 (SE = 0.044), 
larger in magnitude in the narrow window and approaches significance at 10% level 
(p = 0.13). The fact that the coefficient is slightly smaller in the full sample suggests 
the longer window dilutes the association with noise from periods further from the policy.

## 4.3 Alternative functional form (Column 6)

**Did the result hold?** The levels specification uses MWh directly rather
than log(MWh), so the coefficient is not directly comparable in magnitude.
What matters is the **sign** and **significance**, if both are consistent
with the preferred specification, the log transformation is not driving the
result. In levels, the Post × Coal Share coefficient is -852.119 (SE = 461.668), 
which is insignificant, consistent with the log specification. However the standard 
error is enormous relative to the coefficient, and the temperature coefficients also 
become very large and unstable. This suggests the levels specification is poorly behaved, 
likely because the raw MWh values are heavily influenced by the scale differences across 
states that the log transformation was designed to address. The log specification is 
clearly more appropriate for this data.

## 4.4 Alternative inference (Column 7-8)

**Did the result hold?** Compare the SE on `Post × Coal Share` between
the clustered preferred and the HC3 version.

If the HC3 SE is smaller than the clustered SE, it means clustering was
actually making inference more conservative than necessary, and the result
might look stronger under HC3. If the HC3 SE is larger, clustering was
understating the uncertainty even further than the small-cluster problem
suggests. Under HC3 robust SEs, the Post × Coal Share coefficient is identical to the 
preferred specification (-0.06). Meanwhile, the HC3 SE of 0.038 is smaller than the 
clustered SE of 0.054. This means clustering was actually producing more conservative 
standard errors than HC3. However, under HC3 the result is closer to significance 
(p ≈ 0.116). This is noteworthy as it suggests the preferred specification with clustered 
SEs is being conservative, and the result is somewhat more precise under the alternative 
inference approach.

To further probe the significance found in the Add Coal Level specification (Column 3), 
the combined check applies HC3 robust standard errors to the same specification. Recall 
that the HC3 SE is smaller than the clustered SE in this dataset (0.038 vs 0.054 in the 
preferred spec), reflecting that between-state heterogeneity rather than within-state 
serial correlation is the dominant feature of the error structure. When HC3 is applied 
to the Add Coal Level specification, the already-significant result strengthens. The 
coefficient remains at −0.098 but significance improves from the 10% level (clustered SE 
= 0.045) to the 5% level (HC3 SE = 0.043). This confirms that the significance in the Add 
Coal Level specification is not inflated by the clustered SE being too conservative, 
if anything, the clustered SE was understating precision relative to HC3. The result 
becomes more significant, not less, when the more appropriate small-sample SE correction 
is applied.

## 4.5 Placebo test (Column 9)

**Did the result hold?** If the placebo coefficient (`Placebo × Coal Share`) is also 
negative and of similar magnitude, it implies the negative association existed before 
the policy was introduced. It means the main result is likely capturing a pre-existing 
trend rather than anything specific to the carbon pricing period. If the placebo
coefficient is close to zero or positive, it provides some reassurance that
the negative association in the main result is at least specific to the post-July 2012 
period. The placebo coefficient is -0.011 (SE = 0.025), which is close to zero and 
insignificant. This support the credibility of the main finding as the negative association 
does not appear before the policy, suggesting teh effect appear specific to the 
post-policy period.

---

# 5. Overall Assessment

Across eight robustness checks, the Post × Coal Share coefficient is consistently 
negative, ranging from −0.033 to −0.098 across log specifications. The direction 
never flips sign. Statistical significance emerges in two specifications: when coal share is 
included as a standalone level term (Column 3, significant at 10%) and when this is combined 
with HC3 robust standard errors (Column 8, significant at 5%). The coefficient magnitude of 
−0.098 is identical across both specifications, confirming that the point estimate is driven 
entirely by adding coal share as its own variable and the SE correction does not change what 
the model estimates. However, the SE correction does strengthen the significance level, applying 
HC3 reduces the SE slightly from 0.045 to 0.043, pushing significance from the 10% to the 5% 
level. Together this means: the negative association of −0.098 comes from separating coal share 
from the state fixed effects, and the HC3 correction confirms this estimate is precisely enough estimated 
to clear the 5% significance threshold. The preferred specification's insignificance is therefore 
a consequence of near-collinearity between coal share and state fixed effects rather than a 
true absence of association. The placebo coefficient of −0.011 remains near zero and insignificant, 
confirming the negative association is specific to the post-policy period. 

The specifications that do not reach significance (no controls, drop TAS, narrow window, and HC3 on 
the preferred spec) do not undermine the main finding but rather reflect the statistical fragility of 
working with a small panel of five states and 100 observations. In each case the direction is preserved 
and the coefficient remains economically meaningful in magnitude. The levels specification fails entirely 
due to scale instability, confirming the log transformation is the appropriate functional form rather than 
casting doubt on the association itself. Together, the surviving and failing patterns suggest the main 
finding is credible in direction and economically meaningful in magnitude, but statistically sensitive to 
sample size and specification choices. This sensitivity is inherent to the data, rather than a reflection 
of a weak underlying association. The result should still be interpreted as descriptive and subject to the 
limitations noted in the primary analysis.